---
title: "Data Manipulation Tutorial"
author: Puxin
date: "28 Aug 2016"
output: html_document
---

tidyr & lubridate & stringr & dplyr & data.table & some tricks

## _tidyr_

```{r}
#setwd("/home/creatrol/ws/R/Tutorials")
#library(devtools)
#devtools::install_github("hadley/tibble")
#devtools::install_github("hadley/dplyr")
#devtools::install_github("hadley/tidyr")
library(dplyr)
library(tidyr)
```

### Gather columns into key-value pairs
```{r}
wide_df <- data.frame(col = c("X_a","y_b"),
                      A = c(1,2), b = c(2,5), C= c(3,6))
wide_df
(long_df <- gather(wide_df, my_key, my_val, - col))
```

### Spread key-value pairs into columns
```{r}
spread(long_df, my_key, my_val)
```

### Separate one column intp multiple

* 4th arguement (... ,sep = "_")
```{r}
(separate_df <- separate(wide_df, col, c("upper", "lower")))
```

### Unite multiple columns into one

* 4th arguement (... ,sep = "_")
```{r}
(unite_df <- unite(separate_df, col, upper, lower, sep = "-"))
```

### arrange
```{r}
arrange(separate_df, lower)
arrange(separate_df, desc(lower))
```

### Some Issue in practical case
```{r}
# first  --  To handle the problem with "." can not be as sep
iris.tidy <- gather(iris, Part.Measure, Value, -Species)
### way1
iris.tidy2 <- separate(iris.tidy, Part.Measure, c("Part", "Measure"), sep = "\\.")
### way2
iris.tidy2 <- iris.tidy %>% 
  mutate(Part.Measure = gsub(pattern = "\\.", replacement = "_", x = Part.Measure)) %>% 
  separate(Part.Measure, into = c("Part", "Measure"), sep= "_")
# second  --  To handle the error: Duplicate identifiers for rows
iris.tidy2$row <- 1:nrow(iris.tidy2)
iris.tidy <- spread(iris.tidy2, Species, Value)
iris.tidy2 <- iris.tidy2[,1:4]
```

---

## _lubridate_

* date & time

```{r}
library(lubridate)
```

### Basic functions
```{r}
ymd("2016-08-31")
ymd("2016 August 31")
mdy("August 31, 2016")
hms("13:30:09")
ymd_hms("2016/08/31 13.30.09")
```

---

## _stringr_

```{r}
library(stringr)
```

### key functions
```{r}
# Trim leading and trailing white space
str_trim("    this is a    test   ")
# Pad string with x
str_pad("22343", width = 7, side = "left", pad = "x")
# Create character vector of names
friend <- c("Sarah", "Tom", "Tony")
# Search for string in vector
str_detect(friend, "Tony")
# Replace string in vector
str_replace(friend, "Tony", "David")
tolower("I am Not TALKING BOY!!")
toupper("i am a good one..")
```

----

## _dplyr_

```{r}
#setwd("/home/creatrol/ws/R/Tutorials")
library(dplyr)
chicago <- readRDS("chicago.rds")
# To show all columns
options(dplyr.width = Inf)
```

### tbl

* Convert data.frame to table

```{r}
chicagoTable <- tbl_df(chicago)
class(chicago)
class(chicagoTable)
names(chicago)
names(chicagoTable)
cities <- chicagoTable$city
head(cities)
summary(chicago)
summary(chicagoTable)
chicagoTable
str(chicago)
glimpse(chicagoTable)
unique(chicagoTable$city)
chicagoTable[1,1]
```

* change labels

```{r}
changeName <- c("chic" = "chicago", "ny" = "newYork")
chicagoTable$cityName <- changeName[chicagoTable$city]
glimpse(chicagoTable)
chicago$cityName <- changeName[chicago$city]
glimpse(chicago)
```

### Select (column)

* starts_with("X"): every name that starts with "X",
* ends_with("X"): every name that ends with "X",
* contains("X"): every name that contains "X",
* matches("X"): every name that matches "X", where "X" can be a regular expression,
* num_range("x", 1:5): the variables named x01, x02, x03, x04 and x05,
* one_of(x): every name that appears in x, which should be a character vector.

```{r}
chicago <- readRDS("chicago.rds")
dim(chicago)
head(select(chicago,1:5))
names(chicago)[1:3]
head(select(chicago,city:dptp))
head(select(chicagoTable,city:dptp))
head(select(chicago,1:3, pm25tmean2),10)
head(select(chicago, -(city:dptp)))
#################################### Equivalent base R
i <- match("city", names(chicago))
j <- match("dptp", names(chicago))
head(chicago[,-(i:j)])
####################################
head(select(chicago, starts_with("d")))
head(select(chicago, ends_with("2")))
head(select(chicago, matches("mean")))
head(select(chicago, contains("mean")))
```

### Filter (row)

```{r}
chi.f <- filter(chicago, pm25tmean2 > 30)
head(select(chi.f, 1:3, pm25tmean2), 10)
chi.f <- filter(chicagoTable, pm25tmean2 > 30)
head(select(chi.f, 1:3, pm25tmean2), 10)
chi.f <- filter(chicago, pm25tmean2 > 30 & tmpd > 80)
head(select(chi.f, 1:3, pm25tmean2, tmpd), 10)
```

### Arrange (reorder row)

* Ascending

```{r}
chicago <- arrange(chicago,date)
head(select(chicago,date,pm25tmean2),3)
tail(select(chicago,date,pm25tmean2),3)
chicagoTable <- arrange(chicagoTable,date)
head(select(chicagoTable,date,pm25tmean2),3)
head(select(chicagoTable,date,pm25tmean2),3)
```

* Descending

```{r}
chicago <- arrange(chicago, desc(date))
head(select(chicago,date,pm25tmean2),3)
tail(select(chicago,date,pm25tmean2),3)
```

### Rename (variables / column)

```{r}
head(chicago[,1:5],3)
chicago <- rename(chicago, dewpoint = dptp,
                  pm25 = pm25tmean2)
head(chicago[,1:5],3)
```

### Mutate

```{r}
chicago <- mutate(chicago,
                  pm25detrend = pm25 - mean(pm25, na.rm = T))
head(select(chicago, pm25, pm25detrend))
tail(select(chicago, pm25, pm25detrend))
```

* ifelse

```{r}
chicago <- mutate(chicago, 
                  tempcatifelse = ifelse(tmpd <= 80, "cold", "hot"))
head(chicago,3)
```

### Group_by (statistics summarize) summarize

* One sample

```{r}
head(chicago,3)
chicago <- mutate(chicago,
                  tempcat = factor(1 * (tmpd > 80),
                            labels = c("cold","hot")))
head(chicago,3)
hotcold <- group_by(chicago, tempcat)
class(hotcold)
summarize(hotcold, pm25 = mean(pm25, na.rm = T),
          o3 = max(o3tmean2),
          no2 = median(no2tmean2))
```

#### Aggregate functions

* min(x) - minimum value of vector x.
* max(x) - maximum value of vector x.
* mean(x) - mean value of vector x.
* median(x) - median value of vector x.
* quantile(x, p) - pth quantile of vector x.
* sd(x) - standard deviation of vector x.
* var(x) - variance of vector x.
* IQR(x) - Inter Quartile Range (IQR) of vector x.
* diff(range(x)) - total range of vector x.

#### dplyr aggregate functions

* first(x) - The first element of vector x.
* last(x) - The last element of vector x.
* nth(x, n) - The nth element of vector x.
* n() - The number of rows in the data.frame or group of observations that summarise() describes.
* n_distinct(x) - The number of unique values in vector x.

```{r}
chicago <- readRDS("chicago.rds")
head(chicago)
summarise(chicago, num = nth(tmpd, 2))
summarise(chicago, uniqueNumber = n_distinct(city))
```

### %>% the pipe operator

```{r}
chicago %>% mutate(month = as.POSIXlt(date)$mon + 1) %>% group_by(month) %>% summarize(pm25 = mean(pm25tmean2, na.rm = T),
          o3 = max(o3tmean2, na.rm = T),
          no2 = median(no2tmean2, na.rm = T))
```

### dplyr and mySQL databases
```{r, eval=FALSE}
# Set up a connection to the mysql database
my_db <- src_mysql(dbname = "dplyr", 
                   host = "courses.csrrinzqubik.us-east-1.rds.amazonaws.com", 
                   port = 3306, 
                   user = "student",
                   password = "datacamp")

# Reference a table within that source: nycflights
nycflights <- tbl(my_db, "dplyr")
```


---

## _data.table_

* the structure of data.table is DT[i,j,by]

### Create and Subset (filter row i)
```{r}
library(data.table)
DT <- data.table(x = c("A", "B", "C"), y = 1:9, z = (2:10) *2)
DT[3:4]                                         # no need ","
DT[.N]                                          # the last row
```

### Select / Summarize Columns and Add function (j)
```{r}
# select and mutate in dplyr (a little diff)
DT[,.(x, y, diff = z-y, Ymean = mean(y))]
# summarize in dplyr
DT[,.(Zmean = mean(z), Yvar = var(y))]
# Insert functions
DT[, plot(y,z)]
DT[, {print(x)
      hist(z)
      NULL   }]
```

### Doing (j) by Group
```{r}
DT[, .N, by = .(x)]                                 # the number of rows in this group
DT[,.(Ymean = mean(y)), by = .(x)]
DT[, .(y, Zmean = mean(z)), by = .(Grp = y %% 2)]
DT[, .(Zmean = mean(z)), by = .(Grp = y %% 2)]
DT[, mean(z), by = y %% 2]
DT[2:7, mean(z), by = y %% 2]
```

### Chaining
```{r}
ans <- DT[2:9, mean(z), by = x]
ans[order(x)]
####################################### by Chaining
ans <- DT[2:9, mean(z), by = x][order(x)]
ans
``` 

### Subset of Data (.SD) - apply same function to columns (summarize)
```{r}
DT[, lapply(.SD, median), by = x]
DT[, lapply(.SD, median), .SDcols = 2:3]
DT[, lapply(.SD, median), .SDcols = 2:3, by = x]
DT[, lapply(.SD, median), by = x, .SDcols = c("y","z")]
DT[, .(lapply(.SD, median)), by = x, .SDcols = c("y","z")]
DT[, .(lapply(.SD, median), n = .N), by = .(x, y>5), .SDcols = c("y","z")]
```

### Using := in j
```{r}
# Add and update columns (mutate in dplyr)
DT[, c("j","k") := .(rev(y),10:2)];DT
# Remove the columns
DT[, j := NULL];DT
DT[, c("k") := NULL];DT
# Other way to remove
DT[, c("j","k") := .(rev(y),10:2)]
Mycols <- c("j", "k")
DT[, (Mycols) := NULL];DT
# Functional
DT[, `:=` (j = rev(y), k = 10:2)];DT
# by Group
DT[, c := sum(y), by = x];DT
```

### Using set()
```{r}
for (i in 1:9) DT[, set := i+1]
DT
###################### by set(DT, index, column, value)
for (i in 1:9) set(DT, i , 7L, i + 1)
DT
# rename column
setnames(DT, "y", "yrename")
names(DT)
# order columns
setcolorder(DT, c("set", "x", "yrename", "k", "j", "c", "z"))
names(DT)
```

### Indexing (filter)
```{r}
DT[x == "A"]
DT[x %in% c("A", "B")]
# Another way
(w <- DT[, x == "A"])
DT[w]
```

### Keys
```{r}
# create and use
setkey(DT,x)
DT
DT["A"]
# mult
DT["A", mult = "first"]
DT["A", mult = "last"]
# nomatch
DT[c("A", "D")]
DT[c("A", "D"), nomatch = NA] # default
DT[c("A", "D"), nomatch = 0]
# a two-column key
setkey(DT, x, z)
DT["A"]
DT[.("A", 10)]
```

### Rolling Joins
```{r}
DT <- DT[, .(x,k,z)]
DT
setkey(DT, x, k)
DT[.("B")]
DT[.("B", 5)]
# Ordered joins
DT[.("B", 5), roll = TRUE]        # forwards by the second key
DT[.("B", 5), roll = "nearest"]   # nearest by the second key
# Forwards and backwards
DT[.("B", 5), roll = +Inf]        # always forward
DT[.("B", 5), roll = -Inf]        # always backword
# Limited saleness
DT[.("B", 8), roll = 1]          # forward with limited steps
DT[.("B", 8), roll = 2]
DT[.("B", 7), roll = -2]         # backward with limited steps
DT[.("B", 7), roll = -1]
# End control
DT[.("B", 11:12)]
DT[.("B", 11:12), roll = TRUE]
DT[.("B", 11:12), roll = TRUE, rollends = FALSE]
```

## Tricks

* missing and outliers

### Missing values

```{r}
df <- data.frame(A = c(1,NA,8,NA),
                 B = c(3,NA,88,23),
                 C = c(2,45,3,1))
is.na(df)
# Check if there is a NA
any(is.na(df))
# Count the # of NA
sum(is.na(df))
# summary
summary(df)
# Find rows with no NA
complete.cases(df) 
# Another way to remove rows with NAs
na.omit(df)
```

### Outliers and obvious errors
```{r}
set.seed(10)
x <- c(rnorm(30, mean = 15, sd = 5), -5, 28, 35)
# View a boxplot
boxplot(x, horizontal = TRUE)
# Create another dataset
df2 <- data.frame(A = rnorm(100,50,10),
                  B = c(rnorm(99,50,10), 500),
                  C = c(rnorm(99,50,10), -1))
# View summary
summary(df2)
# View histogram
hist(df2$B, breaks = 20)
# View boxplot
boxplot(df2)
```
