---
title: "Statistics"
author: "Puxin"
date: "August 30, 2016"
output: html_document
---

## Basic Statistics

### Histograms and Distributions

#### Looking at distributions by using histograms
```{r, fig.height=3, fig.width=4}
library(Hmisc)
hist(iris$Sepal.Width, breaks = 10)
describe(iris)
```

* Some distribution not normal, for example:
        + Bimodal
        + Positively skewed (left)
        + Negatively skewed (right)
        + Uniform (platykurtic)
        + High peak (leptokurtic)

### Scales of Measurement 

#### Z-scale / z-score

* scale()
```{r, fig.height=3, fig.width=4}
z.Sepal.Width = (iris$Sepal.Width - mean(iris$Sepal.Width)) /sd(iris$Sepal.Width)
hist(z.Sepal.Width, breaks = 10)
###
hist(scale(iris$Sepal.Width), breaks = 10)
```


### Measures of central tendency
```{r}
# mean
mean(iris$Sepal.Width)
#median
median(iris$Sepal.Width)
#mode
Mode <- function(x) {
  ux <- unique(x)
  ux[which.max(tabulate(match(x, ux)))]
}
Mode(iris$Sepal.Width)
```

### Measures of variability
```{r}
var(iris$Sepal.Width)
sd(iris$Sepal.Width)^2
```

## Statistical tests

* z-test
* t-test (simple sample)
* t-test (dependent)
* t-test (independent)


#### plot t-distribution
```{r}
# Generate a vector of 100 values between -4 and 4
x <- seq(-4,4,length = 100)
# Simulate the t-distribution
y_1 <- dt(x, df = 4)
y_2 <- dt(x, df = 6)
y_3 <- dt(x, df = 8)
y_4 <- dt(x, df = 10)
y_5 <- dt(x, df = 12)
# Plot the t-distributions
plot(x, y_1, type = "l", lwd = 2, xlab = "t-value", ylab = "Density", 
     main = "Comparison of t-distributions", col = "black")
lines(x, y_2, col = "red")
lines(x, y_3, col = "orange")
lines(x, y_4, col = "green")
lines(x, y_5, col = "blue")
# Add a legend
legend("topright", c("df = 4", "df = 6", "df = 8", "df = 10", "df = 12"), 
       col = c("black", "red", "orange", "green", "blue"), 
       title = "t-distributions", lty = 1)
```









